Handling missing data in smartphone location logs

### 2017-11-25 thesisReportApa.Rmd
edited a bit and added new APA format

---
title: 'Handling missing data in smartphone location logs'
author: "Boaz Sobrado"
date: "`r format(Sys.time(), '%d %B %Y')`"
bibliography:
  - thesisGPS.bib
  - rReferences.bib
output:
  rmarkdown::pdf_document:
    citation_package: natbib
    fig_caption: yes        
    includes:  
      in_header: tp2.tex

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(ggthemes)
library(ggplot2)
library(knitr)
```

# Abstract
*Using objective location data to infer the mobility measures of individuals is highly desirable, but methodologically difficult. Using commercially gathered location logs from smartphones holds great promise, as they have already been gathered, often span years and can be associated to individuals. However, due to technical constraints this data is more sparse and inaccurate than that produced by specialised equipment. In this paper we present a model which leverages the periodicity of human mobility in order to impute missing data values. Moreover, we will assess the performance of the model relative to currently used methods, such as linear interpolation.*

# Introduction

@@

In most of these studies participants are given a specialised device, resulting in accurate mobility GPS data (*specialised logs*). However, @barnett_inferring_2016 point out that these studies are not scalable due to cost and burden to participants. Moreover, they may be biased because of the introduction of a new device to the participant's life. Because of these drawbacks, specialised logs usually span a short amount of time. @barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*). Another solution is to take advantage of existing smartphone location logs, such as Google Location History, which store location information of millions of users spanning several years [@location_history_timeline_nodate] (*secondary logs*). By law, logs can be accessed and shared by users for free [*European Comission Reference*]. Yet, because they were created for non-academic purposes under engineering constraints, the sensors do not monitor continuously and the resulting logs can be sparse and inaccurate. Hence, two important challenges are dealing with measurement noise and missing data (*messy data*).

@@

## Current methods for imputing & filtering spatiotemporal data

As we have mentioned before, @jankowska_framework_2015 point out that there is little transparency regarding decisions of how to deal with missing data in GPS mobility logs. We suspect this is because the highly interdisciplinary nature of the problem means researchers are unaware of potential solutions. For this reason, we consider it important to briefly discuss a few methods one could consider in order to deal with measurement inaccuracy and missing data problems in spatiotemporal data. In doing so, we will argue that state space models (SSMs) and spatiotemporal imputation methods, which are extensively used to model missing and noisy spatiotemporal data, are not well suited to deal with messy human mobility patterns. Moreover, we discuss in detail two approaches by @palmius_detecting_2017 and @barnett_inferring_2016 which deal explicitly with missing data in mobility patterns from smartphone GPS logs.

### General Spatiotemporal Methods

@@

Spatiotemporal imputation methods generally assume  fixed measurement stations there are several imputation methods for spatiotemporal measurements. For instance, @feng_cutoff:_2014 illustrate their CUTOFF method, which relies on estimating missing values using the nearest observed neighbours in time, using rainfall data from dozens of gauging stations across Australia. Similarly, @zhang_application_2017 use a variety of machine learning methods to present their model based on underground water data in China.

While @feng_cutoff:_2014 claim their model could be used to establish mobility patterns, ostensibly by dividing the sample space into rasters analogous to measurement stations indicating a probability of the individual being there, this seems to be computationally unfeasible. To our knowledge such models have not been implemented.

### Filtering & Mean imputation

@palmius_detecting_2017 deal with measuremement inaccuracy of $D$ by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$. Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter. Moreover, @palmius_detecting_2017 explain:

@@

### Barnett's model

@barnett_inferring_2016 deal with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes. @barnett_inferring_2016 handle missing data by:

@@

This method can be extended to imputing the data based on temporally, spatially or periodically close flights and pauses. In other words, for a given missing period, the individual's mobility can be estimated based on  measured movements in that area, at that point in time or movements in the last 24 hours.

This paper is to the best of our knowledge the only attempt at establishing a principled approach to this problem.

*perhaps remove the section on alternative for the december deadline*
### Alternative models

Alternatives to state space models include long range-persistence models, such as cascading walks models  and the FarOut model which rely on self-similarity and autoregressive characteristics [@han_cascading_2015; @sadilek_far_2016]. The latter uses Fourier analysis and PCA to extract cyclical patterns in an individual's behaviour and reduce the dimensionality of the extracted features and yields interpretable predictions for an individuals location months in advance.

@@

The data used to train the imputation methods was collected between 2013 and 2017 on different Android devices from several individuals. The table below provides more details:

@@

Analyses were performed using R and a multitude of other packages [@ggplot2;@dplyr;@leaflet; @ggthemes;@base;@sp1;@sp2].

### Data pre-processing

Three different filtering methods were tested. First, the filtered rolling-median downsampling method described by @palmius_detecting_2017. The second method is a weighted mean approach taking $f(a)$ as a weight. Finally, an extended Kalman filter. The output of all of these methods was taken as the input of the imputation methods.

@@

Four imputation methods were selected in order to cover techniques applied in the literature. Briefly, the mean imputation method described by @palmius_detecting_2017, the model developed by @barnett_inferring_2016 as well as an adapted FarOut model [@sadilek_far_2016]. Simple linear interpolation was used as a benchmark model.

@@

The entire length of the secondary logs were used as a training set. The specialised logs were used as a test set. The missing data imputation models were evaluated both directly, and on on two computed measures: amount of trips made and distance traveled.

@@

The evaluation on computed measures involved calculating a mobility trace following the rectangual method of @rhee_human_2007 for each imputed dataset. Like @barnett_inferring_2016 we calculate bias by substracting the estimated measure under each approach for the same measure calculated on the full data. For simulation-based imputation approaches a mean value over 100 samples was taken.

## Results

## Discussion

## References

>>>>

---
title             : "Handling missing data in smartphone location logs"
shorttitle        : "Missing Data in Smartphone Location Logs"

author: 
  - name          : "Boaz Sobrado"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "boaz@boazsobrado.com"

affiliation:
  - id            : "1"
    institution   : "Utrecht University"

author_note: |
  Department of Methodology & Statistics

  Submitted as a research report.

abstract: |
  Using objective location data to infer the mobility measures of individuals is highly desirable, but methodologically difficult. Using commercially gathered location logs from smartphones holds great promise, as they have already been gathered, often span years and can be associated to individuals. However, due to technical constraints this data is more sparse and inaccurate than that produced by specialised equipment. In this paper we present a model which leverages the periodicity of human mobility in order to impute missing data values. Moreover, we will assess the performance of the model relative to currently used methods, such as linear interpolation.

keywords          : "GPS, Human Mobility"
wordcount         : "X"

bibliography:
  - thesisGPS.bib
  - rReferences.bib

figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes

lang              : "english"
class             : "man"
output            : papaja::apa6_pdf
citation_package: natbib
---

```{r load_packages, include = FALSE}
library("papaja")
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(ggthemes)
library(ggplot2)
library(knitr)
```

@@

In most of these studies participants are given a specialised device, resulting in accurate mobility GPS data (*specialised logs*). However, @barnett_inferring_2016 point out that these studies are not scalable due to cost and burden to participants. Moreover, they may be biased because of the introduction of a new device to the participant's life. Because of these drawbacks, specialised logs usually span a short amount of time. @barnett_inferring_2016 advocate installing a custom-made tracking app on user's phones (*custom logs*). Another solution is to take advantage of existing smartphone location logs, such as Google Location History, which store location information of millions of users spanning several years [@location_history_timeline_2017] (*secondary logs*). By law, logs can be accessed and shared by users for free [@commission_protecting_2017]. Yet, because they were created for non-academic purposes under engineering constraints, the sensors do not monitor continuously and the resulting logs can be sparse and inaccurate. Hence, two important challenges are dealing with measurement noise and missing data.

@@

## Current methods for imputing missing spatiotemporal data

As we have mentioned before, @jankowska_framework_2015 point out that there is little transparency regarding decisions of how to deal with missing data in GPS mobility logs. We suspect this is because the highly interdisciplinary nature of the problem means researchers are unaware of potential solutions. For this reason, we consider it important to briefly discuss a few methods one could consider in order to deal with measurement inaccuracy and missing data problems in spatiotemporal data. In doing so, we will argue that state space models (SSMs) and spatiotemporal imputation methods, which are extensively used to model missing and noisy spatiotemporal data, are not well suited to deal with human mobility patterns. Moreover, we discuss in detail two approaches by @palmius_detecting_2017 and @barnett_inferring_2016 which deal explicitly with missing data in mobility patterns from smartphone GPS logs.

@@

In addition to SSMs there are spatiotemporal imputation methods often used in climate or geological research for estimating missing data. For instance, @feng_cutoff:_2014 illustrate their CUTOFF method, which relies on estimating missing values using the nearest observed neighbours in time, using rainfall data from dozens of gauging stations across Australia. Similarly, @zhang_application_2017 use a variety of machine learning methods to present their model based on underground water data in China.

The difficulty of applying this class of spatiotemporal imputation models to human mobility tracks lies in the fact that these models generally assume fixed measurement stations (such as rainfall gauging stations). While @feng_cutoff:_2014 claim their model could be used to establish mobility patterns, ostensibly by dividing the sample space into rasters analogous to measurement stations indicating a probability of the individual being there, this seems to be computationally unfeasible. To our knowledge such models have not been implemented for mobility traces.

On the other hand, a few researchers have explicitly attempted to impute missing data from human mobility patterns. @palmius_detecting_2017 deal with the measuremement inaccuracy of $D$ in custom logs by removing from the data set all unique low-accuracy $a$ data points that had  $\frac{d}{dt}D > 100 \frac{km}{h}$. Subsequently the researchers down sample the data to a sample rate of 12 per hour using a median filter. Moreover, @palmius_detecting_2017 explain:

@@

@barnett_inferring_2016 follow a different approach which is, to the best of our knowledge, the only pricipled approach to dealing with missing data in human mobility data. @barnett_inferring_2016 work with custom logs where location is measured for 2 minutes and subsequently not measured for 10 minutes. In the words of the authors, @barnett_inferring_2016 handle missing data by:

@@

This method can be extended to imputing the data based on temporally, spatially or periodically close flights and pauses. In other words, for a given missing period, the individual's mobility can be estimated based on measured movements in that area, at that point in time or movements in the last 24 hours (*circadian proximity*).

@@

The data used to train the imputation methods was collected between 2013 and 2017 on different Android devices from several individuals (table 1).

@@

Analyses were performed using R and a multitude of other statistical packages [@ggplot2;@dplyr;@leaflet;@base;@sp1;@sp2].

### Data pre-processing & filtering

The goal of filtering was to remove noise from the measurements and to aggregate multiple measurements into 12 per hour. Three different filtering methods were tested:

1. The filtered rolling-median downsampling method described by @palmius_detecting_2017.
2. A weighted mean approach taking $f(a)$ as a weight.
3. A Kalman filter commonly used for GPS measurements [@doust_smoothing_2013].

The output of all of these methods was taken as the input of the imputation methods.

@@

Three imputation methods were selected in order to cover a wide range of techniques applied in the literature:

1. Mean imputation as described by @palmius_detecting_2017.
2. The model developed by @barnett_inferring_2016 using spatial, temporal and circadian proximity.
3. Simple linear interpolation was used as a benchmark model.

@@

The entire length of the secondary logs were used as a training set. The specialised logs were used as a test set. The missing data imputation models were evaluated both directly, and on two computed measures: amount of trips made and distance traveled.

@@

The evaluation on computed measures involved calculating a mobility trace following the rectangular method of @rhee_human_2007 for each imputed dataset. Like @barnett_inferring_2016 we calculate bias by substracting the estimated measure under each approach for the same measure calculated on the full data. For simulation-based imputation approaches a mean value over 100 samples was taken.

\newpage

# References

\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}
